A full approximation scheme multilevel method \\ for solving nonlinear variational inequalities

### 2024-07-24 talk-ismp/slides.tex
a bit shorter and better

\only<1>{
\begin{itemize}
\item consider a nonlinear elliptic PDE problem:
	$$F(u) = \ell$$

	\begin{itemize}
	\item[$\circ$] $\mathcal{V}=H^1(\Omega)$, $F : \mathcal{V} \to \mathcal{V}'$ continuous, $\ell\in \mathcal{V}'$
	\item[$\circ$] example is the Liouville-Bratu problem: $-\grad^2 u - e^u = f$
	\end{itemize}
\item discretization\footnote{discretization = finite element (FE) method, throughout this talk} gives algebraic system on fine mesh $\Omega^h$:
    $${\color{FireBrick} F^h(u^h) = \ell^h}$$
\end{itemize}

\vspace{4mm}
}
\only<2>{
\begin{itemize}
\item \emph{goal}: to solve $F^h(u^h) = \ell^h$ on $\Omega^h$
\item suppose $w^h$ is a not-yet-converged iterate:
    $$r^h=\ell^h - F^h(w^h), \qquad \|r^h\| > \text{TOL}$$
\item how can we improve $w^h$ \emph{without} globally linearizing $F^h$?

	\begin{itemize}
	\item are there alternatives to Newton's method?
	\end{itemize}
\item notes:
    \setbeamertemplate{enumerate item}{\emph{\roman{enumi})}}
    \begin{enumerate}
    \item \emph{residual} $r^h = \ell^h - F^h(w^h)$ is computable but \emph{error} $e^h = u^h-w^h$ is unknown
    \item equation can be rewritten as \qquad $F^h(u^h) - F^h(w^h) = r^h$
    \end{enumerate}
\end{itemize}
}
\only<3>{
\begin{itemize}
\item \emph{goal}: from iterate $w^h$, to solve {\color{FireBrick} $F^h(u^h) - F^h(w^h) = r^h$}
\item for $F^h$ linear, one may convert to the error equation
    $$F^h(e^h) = r^h$$
with approximate solution $\tilde e^h$, to improve the iterate
    $$w^h \leftarrow w^h+\tilde e^h$$
\end{itemize}

\vspace{5mm}
}
\only<4>{
\begin{itemize}
\item \emph{updated goal}: use a coarser mesh $\Omega^H$ to somehow estimate the solution $u^h$ in the nonlinear \emph{correction equation}
    $${\color{FireBrick} F^h(u^h) - F^h(w^h) = r^h}$$
\item \emph{smoothers}: there are inexpensive algorithms which improve $w^h$ by efficiently removing high-frequency error components
	\begin{itemize}
	\item sweeping through and solving nodewise problems (nonlinear Gauss-Seidel) is a smoother
	\end{itemize}
\item how about low-frequency error components?
\end{itemize}

\vspace{5mm}
}
\only<5>{
\begin{itemize}
\item Brandt's (1977) \emph{full approximation scheme} (FAS) equation:
	$${\color{FireBrick} F^H(u^H) - F^H(\iR w^h) = R \, r^h(w^h)}$$

@@

\item rewritten FAS equation: let $\ell^H = F^H(\iR w^h) + R\, r^h(w^h)$ then
    $${\color{FireBrick} F^H(u^H) = \ell^H}$$
\end{itemize}
}
\end{frame}

@@

\begin{itemize}
\item FAS multigrid is a fast solver on nice nonlinear PDE problems
\item example: Liouville-Bratu equation
    $$-\nabla^2 u - e^u = 0$$
with Dirichlet boundary conditions on $\Omega=(0,1)^2$
%\item minimal problem-specific code:
%    \begin{enumerate}
%    \item[1.] residual evaluation on grid level: $F^j(\cdot)$
%    \item[2.] pointwise smoother: $\phi_i(c) = 0 \,\forall i$
%        \begin{itemize}
%        \item[$\circ$] nonlinear Gauss-Seidel iteration
%        \end{itemize}
%    \item[3.] coarsest-level solve can be same as smoother
%    \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}{a meaning for ``fast solver''}

\begin{block}{definition} a solver is \emph{optimal} if work is $O(N)$ for $N$ unknowns
\end{block}

\bigskip
\begin{itemize}
\item one can measure work in flops or run-time
\item constant iterations of $O(N)$ method $\implies$ optimal
\end{itemize}
\end{frame}

@@

\begin{frame}{inequality-constrained problem: flowing glacier ice in a climate}

\begin{itemize}
\item glacier = incompressible, viscous fluid driven by gravity
\item in a \emph{climate} which adds or removes ice at a signed rate $a(t,x,y)$
\item over fixed bed topography with elevation $b(x,y)$
    \begin{itemize}
    \item[$\circ$] data $a,b$ is defined on domain $\Omega \subset \RR^2$
    \end{itemize}
\item to find: ice surface elevation $s(t,x,y)$ and velocity $\bu(t,x,y,z)$
\item subject to constraint $\boxed{s\ge b}$
\end{itemize}

\bigskip
\hfill \mbox{\includegraphics[height=0.24\textheight]{../talk-oxford/images/domain-data.png} \quad $\stackrel{?}{\to}$ \quad \includegraphics[height=0.24\textheight]{../talk-oxford/images/domain-velocity.png}}
\end{frame}

@@

\begin{frame}{test case: ice sheet the size of Greenland}

\centering
\includegraphics[width=\textwidth]{../paper/fixfigs/sialev8scene.png}
\end{frame}

\begin{frame}{FASCD: parallel weak scaling}

\begin{itemize}
\item optimality of FMG solver (left)
\item good parallel \emph{weak scaling} (right)
    \begin{itemize}
    \item[$\circ$] each processor owns $641\times 641$ (sub) mesh
    \item[$\circ$] $P=1024$ run had $20481^2=4.1\times 10^8$ unknowns
    \item[] \dots and 88 meter resolution
    \end{itemize}
\end{itemize}

@@

\bigskip
    \begin{itemize}
    \item[$\circ$] what is ``constraint decomposition''?
    \item[$\circ$] unilateral (lower obstacle) version shown, but bilateral or ``box constraints'' are implemented
    \end{itemize}
\end{itemize}
\end{frame}

@@

    \begin{itemize}
    \item[$\circ$] \emph{non}-unique vector space sum
    \item[$\circ$] one may analyze linear multigrid for PDEs via subspace decomposition
    \end{itemize}
\end{itemize}
\end{frame}

@@

\item this iteration is not practical because you must evaluate all residuals on the finest level
\item two techniques make a practical solver:

    \begin{itemize}
    \item[$\circ$] \emph{defect obstacles} on each level  \hfill {\scriptsize Gr\"aser \& Kornhuber (2009)}
    \item[$\circ$] \emph{FAS coarse corrections}  \hfill {\scriptsize Bueler \& Farrell (2024)}
    \end{itemize}
\end{itemize}
\end{frame}

@@

see Bueler \& Farrell (2024) for:
\begin{itemize}
\item generalization to upper and lower obstacles:
    $$\mathcal{K}^J = \{\underline{\psi}^J \le v^J \le \overline{\psi}^J\}$$
\item stopping criteria
    \begin{itemize}
    \item[$\circ$] evaluates CP/KKT conditions
    \end{itemize}
\item FMG cycle
\item details of $O(m_J)$ smoother: a few iterations of projected, reduced-space Newton with a few iterations of CG or GMRES, plus ILU preconditioning, on the linear step equations
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FASCD = new multilevel solver for VI (free-boundary) problems
    \begin{itemize}
    \item[$\circ$] implemented in Python using the Firedrake library
    \item[$\circ$] observed optimality and actually fast on tested problems
    \end{itemize}
\item many things \textbf{{\color{FireBrick} to do}}:
    \begin{itemize}
    \item[$\circ$] implement in C within PETSc
    \item[$\circ$] prove convergence; compare (Reusken, 1988)
    \item[$\circ$] mesh adaptivity, better smoothers, parabolic examples, \dots
    \end{itemize}
\end{itemize}

>>>>

\only<1>{
\begin{itemize}
\item consider a nonlinear elliptic PDE problem:
	$${\color{FireBrick} F(u) = \ell}$$

	\begin{itemize}
	\item[$\circ$] $\mathcal{V}=H^1(\Omega)$, $F : \mathcal{V} \to \mathcal{V}'$ continuous, $\ell\in \mathcal{V}'$
	\end{itemize}
\item discretization\footnote{discretization = finite element (FE) method, throughout this talk} gives algebraic system on fine mesh $\Omega^h$:
    $${\color{FireBrick} F^h(u^h) = \ell^h}$$
\item suppose $w^h$ is a not-yet-converged iterate:
    $$r^h=\ell^h - F^h(w^h), \qquad \|r^h\| > \text{TOL}$$
\end{itemize}
}
\only<2>{
\begin{itemize}
\item how can we improve $w^h$ \emph{without} globally linearizing $F^h$?

	\begin{itemize}
	\item are there alternatives to Newton's method?
	\end{itemize}
\item rewrite the equation as the nonlinear \emph{correction equation}
    $${\color{FireBrick} F^h(u^h) - F^h(w^h) = r^h}$$
\item \underline{for $F^h$ linear}, convert the correction eqn to the \emph{error equation}
    $$F^h(e^h) = r^h,$$
with approximate solution $\tilde e^h$, and update $w^h \leftarrow w^h+\tilde e^h$
\end{itemize}
}
\only<3>{
\begin{itemize}
\item there are fast \emph{smoothers}: inexpensive algorithms which improve $w^h$ by efficiently removing high-frequency error components
	\begin{itemize}
	\item sweeping through and solving nodewise problems (nonlinear Gauss-Seidel) is a smoother
	\end{itemize}
\item how about low-frequency error components?
\item \emph{idea}: use a coarser mesh $\Omega^H$ to estimate $u^h$ in the correction eqn
\end{itemize}

\vspace{5mm}
}
\only<4>{
\begin{itemize}
\item Brandt's (1977) \emph{full approximation scheme} (FAS) equation:
	$${\color{FireBrick} F^H(u^H) - F^H(\iR w^h) = R \, r^h(w^h)}$$

@@

\item let $\ell^H = F^H(\iR w^h) + R\, r^h(w^h)$
\item rewritten FAS equation:
    $${\color{FireBrick} F^H(u^H) = \ell^H}$$
\end{itemize}
}
\end{frame}

@@

\begin{itemize}
\item FAS multigrid is a \emph{fast solver} on nice nonlinear PDE problems
    \begin{itemize}
    \item[$\circ$] example: Liouville-Bratu equation $-\nabla^2 u - e^u = 0$
    \end{itemize}

\bigskip
\item<2> what does ``fast solver'' mean?
\end{itemize}

\begin{block}<2>{definition} a solver is \emph{optimal} if work is $O(N)$ for $N$ unknowns
\end{block}

\begin{itemize}
\item<2> one can measure work in flops or run-time
\item<2> constant iterations of $O(N)$ method $\implies$ optimal
\end{itemize}
\end{frame}

@@

\begin{frame}{the glacier free-boundary problem}

\begin{itemize}
\item a \emph{glacier} is an incompressible, viscous fluid driven by gravity, with mass from a climate which adds or removes ice at a signed rate, over a fixed bed topography with elevation $b$
\item to find: ice surface elevation $s$ (and velocity $\bu$) subject to constraint $\boxed{s\ge b}$
\item test case: ice sheet the size of Greenland

\bigskip
\begin{center}
\includegraphics[width=0.65\textwidth]{../paper/fixfigs/sialev8scene.png}
\end{center}
\end{itemize}
\end{frame}

@@

\begin{frame}{optimality and weak scaling on a glacier problem}

\begin{itemize}
\item optimality of FMG solver (left)
\item good parallel \emph{weak scaling} (right)
    \begin{itemize}
    \item[$\circ$] each processor owns $641\times 641$ (sub) mesh
    \item[$\circ$] $P=1024$ run had $20481^2=4.1\times 10^8$ unknowns and 88 meter resolution
    \end{itemize}
\end{itemize}

@@

\bigskip
\item<2> but what is ``constraint decomposition''?
\end{itemize}
\end{frame}

@@

    \begin{itemize}
    \item[$\circ$] \emph{non}-unique vector space sum
    %\item[$\circ$] one may analyze linear multigrid for PDEs via subspace decomposition
    \end{itemize}
\end{itemize}
\end{frame}

@@

\item \dots but not practical because you evaluate residuals on finest level
\item two techniques make a practical solver (Bueler \& Farrell, 2024):

    \begin{itemize}
    \item[$\circ$] \emph{defect obstacles} on each level  \hfill {\scriptsize Gr\"aser \& Kornhuber (2009)}
    \item[$\circ$] \emph{FAS coarse corrections}  \hfill {\scriptsize Brandt (1977)}
    \end{itemize}
\end{itemize}
\end{frame}

@@

see Bueler \& Farrell (2024) for:
\begin{itemize}
\item generalization to upper and lower obstacles:
    $$\mathcal{K}^J = \{\underline{\psi}^J \le v^J \le \overline{\psi}^J\}$$
\item stopping criteria
    \begin{itemize}
    \item[$\circ$] evaluates CP/KKT conditions
    \end{itemize}
\item FMG cycle
\item details of the $O(m_j)$ smoother used in the results: a few iterations of projected, reduced-space Newton with a few iterations of CG+ICC or GMRES+ILU on the linear step equations
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item FASCD = new multilevel solver for VI (free-boundary) problems
    \begin{itemize}
    \item[$\circ$] implemented in Python using the Firedrake library
    \item[$\circ$] observed optimality, and actually fast on tested problems
    \end{itemize}
\item many things \textbf{{\color{FireBrick} to do}}:
    \begin{itemize}
    \item[$\circ$] implement in C within PETSc
    \item[$\circ$] prove convergence; compare (Reusken, 1988)
    \item[$\circ$] mesh adaptivity, better smoothers, parabolic examples, \dots
    \end{itemize}
\end{itemize}
